---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# rmskProfiler

<!-- badges: start -->
<!-- badges: end -->

This package provides an end-to-end solution for accurately quantifying 
transposable elements from RNA-seq data at the locus-level and importing 
into R for downstream analysis. 

## Installation

You can install the development version of `rmskProfiler` from [GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pak("coriell-research/rmskProfiler")
```

### Salmon dependency

This package assumes that you have a recent version of [Salmon](https://salmon.readthedocs.io/en/latest/salmon.html) installed and 
available on your PATH. If not, please follow the latest [installation](https://salmon.readthedocs.io/en/latest/building.html#binary-releases) 
instructions before using this package.

## Overview of the method

This package simply creates a Salmon index using GENCODE v48 annotated transcripts, unique RepeatMasker repeat regions, and a full genomic decoy and provides utilities for importing Salmon quants into R. The imported counts are stored along with their annotation information (including GenomicRanges) as a SummarizedExperiment object that is ready for downstream analysis. 

By default, not all repeats are included in the generated index. "Simple_repeat", "Low_complexity", "Satellite", "RNA", "rRNA", "snRNA", "scRNA", "srpRNA", "tRNA", and "Unknown" loci are excluded. Additionally, TE loci with lengths <=31 bp are removed (as these are the size of the minimum k-mer length used by Salmon). 

After the index has been created, reads are then quantified using a minimum 
of 30 Gibbs re-samples. The Gibbs re-sampled quants are imported into R using `edgeR::catchSalmon()`. The original counts from Salmon along with the overdispersion corrected counts from `catchSalmon()` are stored as separate 
assays in the resulting `SummarizedExperiment` object. 

All intermediate files and and the Salmon index are stored in the created
resources directory for further inspection and re-use. 

## Usage

The package has three main components:

1. Generate a `Salmon` index using unique RepeatMasker elements + transcripts + genomic decoy
2. Quantify reads with Salmon using Gibbs sampling
3. Import the Salmon quants as a `SummarizedExperiment` for downstream differential expression analysis

A full pipeline for generation of the index, quantification, and importing of 
quants looks like:

```{r, eval=FALSE}
library(rmskProfiler)


# Generate the Salmon index for humans using 12 threads 
# and save to directory named 'hg38-resources'
generateIndex("hg38-resources", species = "Hs", threads = 12)

fq1 <- c("/path/to/sample1.R1.fq.gz", "/path/to/sample2.R1.fq.gz", "/path/to/sample3.R1.fq.gz")
fq2 <- c("/path/to/sample1.R2.fq.gz", "/path/to/sample2.R2.fq.gz", "/path/to/sample3.R2.fq.gz")
sample_names <- c("sample1", "sample2", "sample3")

# Perform quantification with Salmon on fastq files
# NOTE: Once the index is built you can quantify reads using any workflow
#       just make sure you perform Gibbs re-sampling
salmonQuant(
  fq1 = fq1, 
  fq2 = fq2, 
  sample_names = sample_names, 
  resource_dir = "hg38-resources", 
  out_dir = "quants", 
  n_gibbs = 30,
  "--gcBias",       # Additional arguments can be passed as character strings
  "--seqBias",
  "--posBias",
  "--threads 12"
  )

# Import the transcripts and TE loci counts as a SummarizedExperiment object
# rowData contains transcript and TE annotations and GRanges
se <- importQuants("quants", resource_dir = "hg38-resources")

# Proceed to downstream analysis using edgeR
```

### Filtering and aggregating

The imported object contains information on the TE-loci and transcript levels 
but sometimes a gene/subfamily-level analysis is desired. The following code can 
be used to filter out TE-loci assigned to exons and then summarize counts for 
TE-loci to the subfamily level and transcripts to the gene-level. 

Below, any TE-loci that overlaps with an exon, 3'UTR, 5'UTR on either strand or
has more than one exact sequence location in the genome are removed from 
downstream analysis. Protein coding transcripts are selected for downstream 
analysis. 

```{r, eval=FALSE}
library(rmskProfiler)
library(SummarizedExperiment)


# Select loci to keep
loci <- subset(
  rowData(se), 
  !is.na(Hash) & 
  hasUnstrandedExonic == FALSE & 
  hasUnstranded3UTR == FALSE & 
  hasUnstranded5UTR == FALSE &
  N_Loci == 1,
  select = "Hash",
  drop = TRUE
  )

# Select only protein coding transcripts
tx <- subset(
  rowData(se), 
  !is.na(transcript_id) & gene_type == "protein_coding", 
  "transcript_id", 
  drop = TRUE
  )

# Filter the Summarized experiment object
keep <- rownames(se) %in% c(tx, loci)
filtered <- se[keep, ]

# Sum assays to the gene/subfamily level
aggregated <- aggregateCounts(filtered, resource_dir = "hg38-resources", level = "subfamily")
```
